---
title: "Farmer aging and succession"
subtitle: "MtM ABM description"
author: "David O'Sullivan"
date: 19 February 2025
date-modified: 17 October 2025
lightbox: true
filters:
  - nutshell
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: "Contents"
    code-fold: true
    html-math-method: katex
    embed-resources: true
    mermaid:
      theme: neutral
knitr:
  opts_chunk: 
    warning: false
    message: false
    # echo: false
    R.options: 
      knitr.table.html.attr: "quarto-disable-processing=true"
---

## Farmer aging
Farmers age by five years per model tick (specified by hard-coded global `years-per-tick`).

After aging, a farmer is tested for exiting the industry (or retiring or dying, or however we want to think about it). They exit with a probability given by 

$$
p_{\mathrm{exit}} = \left\{
  \begin{array}{cl} 
    \frac{5}{91-\mathrm{age}} & \text{if age} <    91 \\
                            1 & \text{if age} \geq 91
  \end{array}
\right.
$$

This means that farmers are not guaranteed to exit until they are 90, reflecting the surprising numbers of older farmers in survey data.

## Age of farmers at model initialisation
The age of farmers at model initialisation is drawn from a [:beta-binomial distribution](https://en.wikipedia.org/wiki/Beta-binomial_distribution), with $n=70$, $\alpha=5$, and $\beta=4$, and ages then offset by adding 20 (so that 20 is the youngest possible age). An example distribution for 250 farmers is shown below, with the red line showing the associated probability density function.

```{r}
#| echo: false
library(extraDistr)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(1)

get_age_df <- function(num_farmers = 250,
                       age_range = 70,
                       alpha = 5,
                       beta = 4, 
                       min_age = 20) {
  age_draw <- data.frame(Age = 20 + rbbinom(num_farmers, age_range, alpha, beta)) |>
    count(Age)
  data.frame(Age = min_age + 0:age_range) |>
    left_join(age_draw) |>
    mutate(n = if_else(is.na(n), 0, n),
           pdf = dbbinom(0:age_range, age_range, alpha, beta),
           exp_n = pdf * num_farmers) |>
    rename(`Number of farmers` = n)
}

ages_init <- get_age_df()
ggplot(ages_init) +
  geom_col(aes(x = Age, y = `Number of farmers`), fill = "grey") +
  geom_line(aes(x = Age, y = exp_n), colour = "red") +
  theme_minimal()
```

This distribution has been roughly matched to survey evidence of farmer ages (it has not been fitted, just matched approximately by eye).

## Age of new farmers
Incoming farmer ages are also drawn from a beta-binomial distribution with $n=45$, $\alpha=4$, $\alpha=2$ again offset so that the youngest farmers are 20.

```{r}
#| echo: false
ages_succession <- get_age_df(age_range = 45, alpha = 4, beta = 2)
ggplot(ages_succession) +
  geom_col(aes(x = Age, y = `Number of farmers`), fill = "grey") +
  geom_line(aes(x = Age, y = exp_n), colour = "red") +
  theme_minimal()
```

This distribution might appear to skew older than the model age initialisation, but note that the range of values tops out at 65 years as shown in the direct comparison of the probability density functions below.

```{r}
#| echo: false
ages_all <- 
  ages_init |> 
  select(Age, exp_n) |> 
  left_join(ages_succession |> select(Age, exp_n), 
            by = join_by(Age)) |> 
  rename(Initial = exp_n.x, Succession = exp_n.y) |> 
  mutate(Succession = if_else(is.na(Succession), 0, Succession)) |> 
  pivot_longer(c(Initial, Succession))

ggplot(ages_all) +
  geom_ribbon(aes(x = Age, ymax = value, ymin = 0, fill = name, group = name), 
            alpha = 0.5, colour = NA) +
  ylab('Probability density') +
  theme_minimal() +
  theme(legend.title = element_blank())
```

In any case, this paramterisation keeps the farmer age distribution fairly stable over time. 

More detailed fitting and analysis of these (hard-coded) parameter settings around farmer ages is probably desirable.

## Entrant farmers and landuse change
A new farmer will consider whole-farm landuse change (see [Farmer decisions about land use](04-farmer-decisions-land-use.qmd) for details) **under conditions to be determined**. This pathway to land use is more likely to lead to land use change at the whole farm level than the more piecemeal changes that tend to be made by incumbent farmers.

::: {.callout-note collapse="true"}
#### Update History

| Date | Changes
|:-|:------
| 2025-02-19 | Initial post. 
| 2025-10-17 | Updated to reflect changing in aging by 5 years per tick.

:::
